\textsc{Beyond the Iceberg Hypothesis: \\Opening the Black Box of Transport Costs}

### 2018-06-22 redaction/dhp_july2017.tex
changement .tex

\date{February 2018}
 \maketitle
\bigskip

@@

Trade costs are central to international economic analysis. In particular, they are considered a major obstacle to international economic integration and international trade flows. Based on panel data, \citet{novy13} finds  that U.S. trade costs with major trading partners declined on average by about 40 percent between 1970 and 2000. Trade costs declined faster for more developed countries \cite{arvis2016trade}. Still, several papers (mostly based on empirical estimates of the gravity equation) have shown that trade costs still remain a major obstacle to trade (\citealp{Head_Mayer04} or \citealp{Disdier_Head08}, to name a few). Using data over 1989-2000, \citet{anderson_wincoop_jel} thus estimate that average international trade costs for industrialized countries represent a 74\% markup over production costs.

@@

This focus is shared with part of the international trade literature that studies the patterns of transport costs over time, such as \cite{hummels2007} and \cite{Behar_Venables} (among others). Many argue that transport costs have substantially decreased with technological advance in transportation (\citealp{levinson2016box}), infrastructure development and new communication technologies (\citealp{Lafourcade_Thisse}). This view does not make consensus though. Distance as an incompressible cost is still a very important determinant of trade flows (\citealp{Disdier_Head08}); further, progress in transportation techniques (that reduces the constant-dollar cost of transport) should be compared with technical progress in production (that reduces the constant-dollar cost of production) to get a clear picture of the evolution of the transport costs burden. In accordance with this view, \cite{hummels2007} points out that transport costs in proportion of export prices have declined less quickly than expected.\smallskip

According to \cite{hummels2007}, this lower-than-expected decline of transport costs is partly explained by trade composition effects (i.e., changes over time in the baskets of imported goods and/or the origin countries). Based on US imports over 1974-2004, he finds that total transport costs have decreased in both air and maritime shipment, but less so than ``pure'' transport costs (i.e., \textit{ceteris paribus} for given goods, mode, and trade routes) because trade composition effects have partly offset the pure transport costs decrease, thereby attenuating the overall transport costs downward trend.\footnote{Throughout the paper, we refer to ``pure'' or ``per se'' transport costs changes as the changes over time in the transport costs for a given product/country partner, ie excluding trade composition effects.} In this paper, we investigate the robustness of this result. As a major contribution, and in contrast to \cite{hummels2007}, we find that trade composition effects play a minor role in accounting for the trend patterns of transport costs. Rather, the evolution of transport costs ``per se'' constitutes the main driving force behind the overall transport costs decrease, in particular for air transport. In maritime shipment, trade composition effects matter more, but they amplify (rather than offset) the reduction in pure transport costs. In this respect, our results suggest that pure transport costs declined even less than argued by \cite{hummels2007}.\medskip

@@

Second, we find that the main source of the downward trend in overall transport costs (that decreased by roughly 50\% in air shipping and 60\% in vessel shipping over 1974-2013) does come from the reduction of transport cost ``per se''. Trade composition effects only have a limited influence, and when they matter (in maritime transport), they amplify the reduction of pure transport costs, suggesting that pure transport costs declined less than observed costs. This stands in contrast to the conclusion obtained by \cite{hummels2007} on the same database (until 2004). Importantly, we show that the difference of results can be attributed to the new way of modeling of the additive component we offer. Allowing the share of additive costs to vary across time, products and country partners (rather than being constant as in \citealp{hummels2007}), turns out to be key in the decomposition of underlying sources of the decrease of overall transport costs observed over the period. Our result thus deepens the paradox of the less-than-expected decline of transport costs. \smallskip

@@

Our analysis of transportation costs consists in exploiting the difference between commodity-level export and import prices, as in \cite{hummels2007} (among others). The database we use to construct our measure of transport costs comes from US annual "Imports of Merchandise" provided by the Census bureau, spanning from 1974 to 2013. Details on the database are provided in Appendix \ref{app:data}. We first use customs values, quantities and freight costs to recover free-alongside (fas) and cost-insurance-fret (cif) prices, by goods, country of origin and transportation mode.\footnote{The related literature commonly refers to the fob price rather than the fas price. The fas price (for ``Free Alongside Ship'') means that the seller must transport the goods all the way to the dock, close enough to be reached by the crane of the ship it will be transported in. It is also the seller's responsibility to clear the goods for export. The fob price (for ``Free on Board'') means that the seller is obligated to bring the goods all the way to the port, clear the goods for export, and see that they are loaded onto the ship nominated by the buyer. Once the goods clear the railing of the vessel the buyer assumes the risk. Note that this term is used exclusively for maritime and inland waterway transport. While both terms are closely related, the fas price is the one reported by the US Foreign Trade Statistics.} More precisely, the (unit) fas price is computed as the total ``customs value'' in the US trade statistics divided by the shipping weight; in other words, it is the price for one kg of the good net of transportation costs. The cif price is then computed as the sum of the customs value and freight charges, once again divided by the shipping weight. Our dependant variable is finally computed as the ratio of the cif price divided by the fas price. By construction higher than 1, the variable provides a measure of transport costs as a proportion of the good's price, an \emph{ad-valorem} equivalent. This is a quite standard and widespread strategy, as emphasized by \citet{anderson_wincoop_jel}.

@@

\paragraph{Estimation Strategy} We follow \citet{Irrazabal_2015} by considering that \textit{i)} both ad-valorem and additive costs are separable between the origin country ($i$) and the product ($k$) dimensions, and \textit{ii)} this separability is in a multiplicative way for the former and an additive way for the latter. In other words, $\tau_{ik}$ and $t_{ik}$ from Equation (\ref{eq:base_estimee}) are written as:\footnote{Notice that, given the magnitude or order of transport costs, assuming an additive or a multiplicative form for country/product fixed effects does not make a substantial difference since, for small values (as we obtain), we have $\tau_i\times \tau_k -1 \simeq (\tau_i-1) + (\tau_k -1)$ and $t_i+t_k\simeq (1+t_i)\times(1+t_k)-1$.}

@@

After estimating Equation (\ref{eq:model_IetA}), we can re-built a measure of each component, $\widehat{\tau}^{adv}_{is} = \widehat{\tau_{i}} \times \widehat{\tau}_{s(k)}$ for the ad-valorem cost and $\widehat{t}_{is} = \widehat{t}_{i} + \widehat{t}_{s(k)}$ for the additive cost, that are country-sector specific, by year and transport mode. When assuming iceberg costs only (Equation (\ref{eq:model_nlI})), we proceed similarly to get $\widehat{\tau}^{ice}_{is} = \widehat{\tau}_{i} \times \widehat{\tau}_{s}$. In this case, notice that the equation could be estimated relying on a linear form. To preserve comparability of the results, we keep the same non-linear estimation method in both cases though. Similarly as \citet{Irrazabal_2015}, we take the average over the sector-country dimension, using the values of each trade flow ($is$-specific) over total yearly trade as a weighting scheme. We thus recover a ``synthetic estimate'' of each type of transport cost: $\widehat{\tau}^{ice}$ for Model (a), $\widehat{\tau}^{adv}$ and $\widehat{t}$ for Model (b), for each year and transportation mode. These results are reported in Section \ref{sec:results_decomposition}.

@@

Our ambition to re-assess the underlying sources of the trend patterns of international transport costs requires a careful modeling of both the additive and the multiplicative components. The first step of the analysis then consists in providing a quantification of the magnitude of transport costs over time (by transport mode), distinguishing whether the additive component $t_{ik}$ is excluded or included in the estimated equation (Equation (\ref{eq:model_IetA}) or (\ref{eq:model_nlI})). This allows us to provide estimates for the size of both the ad-valorem and the additive components of transport costs, and assess whether additive costs represent a sizeable component of the latter. These results constitute our first original contribution to the literature. In Appendix \ref{app:diagnostic_test}, we provide diagnostic tests about the goodness-of-fit measures, that confirm the empirical relevance of adding the additive component in the estimation.\smallskip

@@

\subsection{Robustness to the weighting scheme \label{sec:robustness}}

As discussed in Appendix \ref{app:comp-effects}, our methodology also differs from \citet{hummels2007} with respect to the weighting scheme used to obtain the evolution of the pure transport costs over time. In this section, we provide a robustness test to this difference of treatment.\smallskip

@@

%\todo{Changer la facon de presenter les resultats, pour harmoniser avec la facon de faire dans le corps du papier. Jerome?}
\begin{table}[htbp]
  \centering
  \caption{Air: Transport costs estimates, 3-digits (selected years)}
\begin{center}
    \begin{tabular}{l|ccccccc}
\hline\hline
Year & 1974  & 1980  & 1990  & 2000  & 2004 & 2010  & 2013   \\
\hline
\multicolumn{8}{l}{\textbf{Model (a) - With only Ad-Valorem TC} ($\widehat{\tau}^{ice}-1$, in \%)}     \\
\hline
Mean  & 6.9& 5.4 &5.0 & 3.6 & 4.0 & 4.2 & 3.4  \\
Median & 5.4 & 3.8 & 4.4 & 2.5 & 2.9 & 3.4 & 2.9  \\
%Standard Error & 0.052 & 0.049 & 0.039 & 0.033 & 0.037 & 0.024 \\
\hline
\multicolumn{8}{l}{\textbf{Model (b) - With Additive \& Ad-Valorem TC}}    \\
\hline
\multicolumn{8}{l}{\textit{Ad-valorem term} ($\widehat{\tau}^{adv}-1$, in \%)}  \\ \hline
Mean & 3.6 & 2.3 & 2.4 &1.7 & 1.9 & 2.6 & 1.7  \\
Median & 2.7 & 1.6 & 1.6 & 1.2 & 1.4 & 2.2 & 1.7 \\
%Standard Error & 0.032 & 0.025 & 0.021 & 0.016 & 0.023 & 0.012  \\

>>>>

\date{April 2018}
 \maketitle
\bigskip

@@

Trade costs are central to international economic analysis. In particular, they are considered a major obstacle to international economic integration and international trade flows. Based on panel data, \citet{novy13} finds  that U.S. trade costs with major trading partners declined on average by about 40 percent between 1970 and 2000. Trade costs declined faster for more developed countries (\citealp{arvis2016trade}). Still, several papers (mostly based on empirical estimates of the gravity equation) have shown that trade costs still remain a major obstacle to trade (\citealp{Head_Mayer04} or \citealp{Disdier_Head08}, to name a few). Using data over 1989-2000, \citet{anderson_wincoop_jel} thus estimate that average international trade costs for industrialized countries represent a 74\% markup over production costs.

@@

This focus is shared with part of the international trade literature that studies the patterns of transport costs over time, such as \cite{hummels2007} and \cite{Behar_Venables}. Many argue that transport costs have substantially decreased with technological advance in transportation (\citealp{levinson2016box}), infrastructure development and new communication technologies (\citealp{Lafourcade_Thisse}). This view does not make consensus though. Distance as an incompressible cost is still a very important determinant of trade flows (\citealp{Disdier_Head08}); further, progress in transportation techniques (that reduces the constant-dollar cost of transport) should be compared with technical progress in production (that reduces the constant-dollar cost of production) to get a clear picture of the evolution of the transport costs burden. In accordance with this view, \cite{hummels2007} points out that transport costs in proportion of export prices have declined less quickly than expected.\smallskip

According to \cite{hummels2007}, this lower-than-expected decline of transport costs is partly explained by trade composition effects (i.e., changes over time in the baskets of imported goods and/or the origin countries). Based on US imports over 1974-2004, he finds that total transport costs have decreased in both air and maritime shipment, but less so than ``pure'' transport costs (i.e., \textit{ceteris paribus} for given goods, mode, and trade routes) because trade composition effects have partly offset the pure transport costs decrease, thereby attenuating the overall transport costs downward trend.\footnote{Throughout the paper, we refer to ``pure'' or ``per se'' transport costs changes as the changes over time in the transport costs for a given product/country partner, ie excluding trade composition effects.} In this paper, we investigate the robustness of this result. As a major contribution, and in contrast to \cite{hummels2007}, we find that trade composition effects play a minor role in accounting for the trend patterns of transport costs. Rather, the evolution of transport costs ``per se'' constitutes the main driving force behind the overall transport costs decrease, in particular for air transport. In maritime shipment, trade composition effects matter more, but they amplify (rather than offset) the reduction in pure transport costs. In this respect, we find that pure transport costs declined even less since the 1970s than argued by \cite{hummels2007}.\medskip

@@

Second, we find that the main source of the downward trend in overall transport costs -by roughly 50\% in air shipping and 60\% in vessel shipping over 1974-2013, does come from the reduction of transport cost ``per se''. Trade composition effects only have a limited influence, and when they matter (in maritime transport), they amplify the reduction of pure transport costs. Otherwise stated, pure transport costs have declined less than observed costs. This stands in contrast to the conclusion obtained by \cite{hummels2007} on the same database (until 2004). Our result thus deepens the paradox of the less-than-expected decline of transport costs. Importantly, we show that the difference of results can be attributed to the new way of modeling of the additive component we offer. Allowing the share of additive costs to vary across time, products and country partners (rather than being constant as in \citealp{hummels2007}), turns out to be key in the decomposition of underlying sources of the decrease of overall transport costs observed over the period. \smallskip

@@

Our analysis of transportation costs consists in exploiting the difference between commodity-level export and import prices, as in \cite{hummels2007}. The database we use to construct our measure of transport costs comes from US annual "Imports of Merchandise" provided by the Census bureau, spanning from 1974 to 2013. Details on the database are provided in Appendix \ref{app:data}. We first use customs values, quantities and freight costs to recover free-alongside (fas) and cost-insurance-fret (cif) prices, by good, country of origin and transportation mode.\footnote{The related literature commonly refers to the fob price rather than the fas price. The fas price (for ``Free Alongside Ship'') means that the seller must transport the goods all the way to the dock, close enough to be reached by the crane of the ship it will be transported in. It is also the seller's responsibility to clear the goods for export. The fob price (for ``Free on Board'') means that the seller is obligated to bring the goods all the way to the port, clear the goods for export, and see that they are loaded onto the ship nominated by the buyer. Once the goods clear the railing of the vessel the buyer assumes the risk. Note that this term is used exclusively for maritime and inland waterway transport. While both terms are closely related, the fas price is the one reported by the US Foreign Trade Statistics.} More precisely, the (unit) fas price is computed as the total ``customs value'' in the US trade statistics divided by the shipping weight; in other words, it is the price for one kg of the good net of transportation costs. The cif price is then computed as the sum of the customs value and freight charges, once again divided by the shipping weight. Our dependant variable is finally computed as the ratio of the cif price divided by the fas price. By construction higher than 1, the variable provides a measure of transport costs as a proportion of the good's price, an \emph{ad-valorem} equivalent. This is a quite standard and widespread strategy, as emphasized by \citet{anderson_wincoop_jel}.

@@

\paragraph{Estimation Strategy} We follow \citet{Irrazabal_2015} by considering that \textit{i)} both ad-valorem and additive costs are separable between the origin country ($i$) and the product ($k$) dimensions, and \textit{ii)} this separability is in a multiplicative way for the former and an additive way for the latter. In other words, $\tau_{ik}$ and $t_{ik}$ from Equation (\ref{eq:base_estimee}) are written as:\footnote{Notice that, given the magnitude or order of transport costs, assuming an additive or a multiplicative form for country/product fixed effects does not make a substantial difference since, for small values (as we obtain), we have $\tau_i\times \tau_k -1 \simeq (\tau_i-1) + (\tau_k -1)$ and $t_i+t_k\simeq (1+t_i)\times(1+t_k)-1$. We also provide a robustness analysis to the separability assumption. Given the much higher number of fixed effects to estimate, we run this robustness check on a sub-sample of goods/origin countries. See Section \ref{sec:robustness} for a detailed presentation of this robustness check.}

@@

After estimating Equation (\ref{eq:model_IetA}), we can re-built a measure of each component, $\widehat{\tau}^{adv}_{is(k)} = \widehat{\tau_{i}} \times \widehat{\tau}_{s(k)}$ for the ad-valorem cost and $\widehat{t}_{is(k)} = \widehat{t}_{i} + \widehat{t}_{s(k)}$ for the additive cost, that are country-sector specific, by year and transport mode. When assuming iceberg costs only (Equation (\ref{eq:model_nlI})), we proceed similarly to get $\widehat{\tau}^{ice}_{is(k)} = \widehat{\tau}_{i} \times \widehat{\tau}_{s(k)}$. In this case, notice that the equation could be estimated relying on a linear form. To preserve comparability of the results, we keep the same non-linear estimation method in both cases though. Similarly as \citet{Irrazabal_2015}, we take the average over the sector-country dimension, using the values of each trade flow ($is$-specific) over total yearly trade as a weighting scheme. We thus recover a ``synthetic estimate'' of each type of transport cost: $\widehat{\tau}^{ice}$ for Model (a), $\widehat{\tau}^{adv}$ and $\widehat{t}$ for Model (b), for each year and transportation mode. These results are reported in Section \ref{sec:results_decomposition}.

@@

Our ambition to re-assess the underlying sources of the trend patterns of international transport costs requires a careful modeling of both the additive and the multiplicative components. The first step of the analysis then consists in providing a quantification of the magnitude of transport costs over time (by transport mode), distinguishing whether the additive component $t_{ik}$ is excluded or included in the estimated equation (Equation (\ref{eq:model_IetA}) or (\ref{eq:model_nlI})). This allows us to provide estimates for the size of both the ad-valorem and the additive components of transport costs, and assess whether additive costs represent a sizeable component of the latter. These results constitute our first original contribution to the literature. We also provide diagnostic tests about the goodness-of-fit measures, that confirm the empirical relevance of adding the additive component in the estimation (see results in In Appendix \ref{app:diagnostic_test}).\smallskip

@@

\section{Robustness analysis \label{sec:robustness}}

\subsection{Robustness to the separability assumption}
In this section, we provide a robustness analysis to the separability assumption of country-good fixed effects, that considers (following so \citealp{Irrazabal_2015}) that both the ad-valorem and the additive costs are separable between the origin country ($i$) and the product ($s(k)$) dimensions. That is, we relax Equations (\ref{eq:ad-valorem}) and (\ref{eq:add}) to rather model $t_{is(k)}$ and $t_{is(k)}$ (on a yearly/transport mode basis). Yet, this comes at the cost of having to estimate a much larger number of fixed effects (for approximatively 200 countries of origin and 200 sectors at the 3-digit level, this means estimating 40,000 versus 400 fixed effects, per year and per transport mode), making the estimation (all the more given our non-linear setting) intractable in practice. This drives us to conduct the analysis on a sub-sample of trade flows, upon which two estimations are conducted (for every year and each transport mode): with the separability assumption (Equations (\ref{eq:ad-valorem}) and (\ref{eq:add}) hold) and without. For each year and transport mode, we select the countries ($i$) and the products (at the 3-digit level, $s(k)$) that constitute 80\% of the total value of trade flows.

Table reports a summary of the results.

\begin{table}[htbp]
  \centering
  \caption{Robustness to the separability assumption}
\begin{center}
    \begin{tabular}{lcccc}
    \hline \hline
    \multicolumn{5}{c}{Mean value over 1974-2013} \\
 \multicolumn{5}{c}{3 digits} \\    \hline \hline
   Mode  & \multicolumn{2}{c}{Air} & \multicolumn{2}{c}{Vessel} \\ \hline
    Additive term ($t/\widetilde{p}$. in \%) &       &       &       &  \\
    Mean  & \multicolumn{1}{c}{1.76} & \multicolumn{1}{c}{1.96} & \multicolumn{1}{c}{2.61} & \multicolumn{1}{c}{2.99} \\
    Median & \multicolumn{1}{c}{0.65} & \multicolumn{1}{c}{0.84} & \multicolumn{1}{c}{1.83} & \multicolumn{1}{c}{2.23} \\
    Ad-valorem term (\$\textbackslash{}tau\$. in \textbackslash{}\%) & \multicolumn{1}{c}{} & \multicolumn{1}{c}{} & \multicolumn{1}{c}{} & \multicolumn{1}{c}{} \\
    Mean  & \multicolumn{1}{c}{1.04} & \multicolumn{1}{c}{0.94} & \multicolumn{1}{c}{2.43} & \multicolumn{1}{c}{2.40} \\
    Median & \multicolumn{1}{c}{0.56} & \multicolumn{1}{c}{0.83} & \multicolumn{1}{c}{2.06} & \multicolumn{1}{c}{2.04} \\
    Diagnostic tests & \multicolumn{1}{c}{} & \multicolumn{1}{c}{} & \multicolumn{1}{c}{} & \multicolumn{1}{c}{} \\
    $R^2$    & \multicolumn{1}{c}{0.80} & \multicolumn{1}{c}{0.68} & \multicolumn{1}{c}{0.67} & \multicolumn{1}{c}{0.45} \\
    SER   & \multicolumn{1}{c}{0.57} & \multicolumn{1}{c}{0.72} & \multicolumn{1}{c}{0.53} & \multicolumn{1}{c}{0.69} \\
    AIC criteria & \multicolumn{1}{c}{4657.60} & \multicolumn{1}{c}{4654.23} & \multicolumn{1}{c}{10744.83} & \multicolumn{1}{c}{10690.80} \\
    Log-likelihood & \multicolumn{1}{c}{-1757.75} & \multicolumn{1}{c}{-2258.24} & \multicolumn{1}{c}{-4018.51} & \multicolumn{1}{c}{-5220.00} \\ \hline
\# observations & \multicolumn{1}{c}{2381} & \multicolumn{1}{c}{2381} & \multicolumn{1}{c}{2798} & \multicolumn{1}{c}{2798} \\
\# origin country & \multicolumn{1}{c}{13.4} & \multicolumn{1}{c}{13.4} & \multicolumn{1}{c}{19.8} & \multicolumn{1}{c}{19.8} \\
\# products & \multicolumn{1}{c}{25.5} & \multicolumn{1}{c}{25.5} & \multicolumn{1}{c}{53} & \multicolumn{1}{c}{53} 
    \end{tabular}%
\end{center}
\label{tab:robustness_separability}
\end{table}

\subsection{Robustness to the weighting scheme }

As discussed in Appendix \ref{app:comp-effects}, one source of difference between our methodology and Hummels's (2007)dwells on the weighting scheme used to obtain the evolution of the pure transport costs over time. In this section, we provide a robustness test to this difference of treatment.\smallskip

@@

%\todo{Changer\begin{table}[htbp]
\begin{table}[htbp]
  \caption{Air: Transport costs estimates, 3-digits (selected years)}
\begin{center}
    \begin{tabular}{l|ccccccc}
\hline\hline 
Year & 1974  & 1980  & 1990  & 2000  & 2004 & 2010  & 2013   \\
\hline
\multicolumn{8}{l}{\textbf{Model (a) - With only Ad-Valorem TC} ($\widehat{\tau}^{ice}-1$, in \%)}     \\
\hline
Mean  & 6.9& 5.4 &5.0 & 3.6 & 4.0 & 4.2 & 3.4  \\
Median & 5.4 & 3.8 & 4.4 & 2.5 & 2.9 & 3.4 & 2.9  \\
%Standard Error & 0.052 & 0.049 & 0.039 & 0.033 & 0.037 & 0.024 \\
\hline
\multicolumn{8}{l}{\textbf{Model (b) - With Additive \& Ad-Valorem TC}}    \\
\hline
\multicolumn{8}{l}{\textit{Ad-valorem term} ($\widehat{\tau}^{adv}-1$, in \%)}  \\ \hline
Mean & 3.6 & 2.3 & 2.4 &1.7 & 1.9 & 2.6 & 1.7  \\
Median & 2.7 & 1.6 & 1.6 & 1.2 & 1.4 & 2.2 & 1.7 \\
%Standard Error & 0.032 & 0.025 & 0.021 & 0.016 & 0.023 & 0.012  \\
